Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-03-02 manuscript/manuscript.tex
removed all latex errors

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$, Christine Lonjou$^{1,2,3}$, Fabienne Lesueur$^{1,2,3}$, GENESIS study collaborators, Dominique Stoppa-Lyonnet$^{4,5,6}$, Nadine Andrieu$^{1,2,3}$, Chloé-Agathe Azencott$^{3,1,2}$}
\affiliation{$^{1}$Institut Curie, PSL Research University, F-75005 Paris, France;\\
  $^{2}$INSERM, U900, F-75005 Paris, France;\\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^{4}$Service de Génétique, Institut Curie, F-75005 Paris, France;\\
  $^{5}$INSERM, U830, F-75005 Paris, France;\\
  $^{6}$Université Paris Descartes;\\
}
\hypersetup{
  pdfauthor={Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, GENESIS investigators, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott},
  pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
  pdfkeywords={},
  pdfsubject={},
  pdflang={English}}
\begin{document}

\onecolumn
\MakeTitle

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks highly associated to this phenotype. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value = $7.8 \times 10^{-5}$ ) and more central than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (\emph{HSPA1A}, \emph{HSPA1B}, and \emph{HSPA1L}) or mitochondrial ribosomes (\emph{MRPS30}, \emph{MRPS31}, \emph{MRPS18B}). We also observed a general dysregulation in the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. 
\end{abstract}

@@

\begin{table}[htbp]
  \caption{\label{tab:method_comparison} Summary of the differences between the studied algorithms.}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Components & Input & Scoring & Reference\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & 1 & Summary & -log\textsubscript{10}(P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \textbf{\emph{Field}}: field in which the algorithm was developed.\\
      \textbf{\emph{Nodes}}: the type of network, either gene (protein-protein interaction network usually) or a SNP network.\\
      \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored.\\
      \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than being dense in high scores and connected.\\
      \textbf{\emph{Components}}: number of connected subnetworks in the solution.\\
      \textbf{\emph{Input}}: genotype data or GWAS summary statistics.\\
      \textbf{\emph{Scoring}}: how SNP/gene P-values are transformed into node scores.\\
      \textbf{\emph{Reference}}: original publication featuring the algorithm.
    \end{tablenotes}
  \end{threeparttable}
\end{table}

@@

\subsection{Network methods successfully identify genes associated with breast cancer}
\label{results:separate_networks}
\begin{table}[htbp]
  \centering
  \begin{threeparttable}
\caption{\label{tab:gene_solutions}
Summary statistics on the results of multiple network methods on the gene-gene interaction network. The first row contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrrrr}
Network & \# genes & \# edges & \mean{Betweenness} & \median{P}\textsubscript{gene} & \(\rho\)\textsubscript{consensus}\\
\hline
HINT HT & 13\,619 & 142\,541 & 16\,706 & 0.46 & 0.066\\
\hline
Consensus & 55 & 117 & 74\,062 & 0.0051 & 1\\
dmGWAS & 194 & 450 & 49\,115 & 0.19 & 0.41\\
heinz & 4 & 3 & 113\,633 & 0.0012 & 0.21\\
HotNet2 & 440 & 374 & 7\,739 & 0.048 & 0.31\\
LEAN & 0 & 0 & - & - & -\\
SConES GI & 0 (1) & 0 & - & - & -\\
SigMod & 142 & 249 & 92\,603 & 0.0083 & 0.73\\
\end{tabular}
\begin{tablenotes}
  \textbf{\emph{\# genes}}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses.\\
  \textbf{\mean{Betweenness}}: mean betweenness of the selected genes in the PPIN.\\
  \textbf{\median{P}\textsubscript{gene}}: median P-value of the selected genes.\\
  \textbf{\(\rho\)\textsubscript{consensus}}: Pearson's correlation between the subnetwork and the consensus network.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

\begin{table}[htbp]
\begin{threeparttable}
\caption{\label{tab:snp_solutions}
Summary statistics on the results of SConES on the three SNP-SNP interaction networks. The first row within each block contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrllr}
Network & SNPs & Edges & Subnetworks & \mean{Betweenness} & \median{P}\textsubscript{SNP}\\
\hline
GS & 197\,083 & 197\,060 & - & 2.03 \texttimes{} 10\textsuperscript{7} & 0.49\\
SConES GS & 1\,590 & 1\,585 & 5 & 2.52 \texttimes{} 10\textsuperscript{7} & 0.023\\
\hline
GM & 197\,083 & 6\,442\,446 & - & 3.99 \texttimes{} 10\textsuperscript{6} & 0.49\\
SConES GM & 1\,692 & 177\,611 & 5 & 4.40 \texttimes{} 10\textsuperscript{6} & 0.055\\
\hline
GI & 197\,083 & 28\,733\,720 & - & 1.46 \texttimes{} 10\textsuperscript{6} & 0.49\\
SConES GI & 408 & 539 & 5 & 9.33 \texttimes{} 10\textsuperscript{6} & 0.076\\
\end{tabular}
\begin{tablenotes}
\mean{Betweenness}: mean betweenness of the selected SNPs in the corresponding full network; \median{P}\textsubscript{SNP}: median P-value of the selected SNPs.
\end{tablenotes}
\end{threeparttable}
\end{table}

\begin{table}[htbp]
\begin{threeparttable}
  \caption{\label{tab:scones_gene_solutions}
Summary statistics on the results of multiple network methods on the gene-gene interaction network. The first row contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrrlr}
Network & Genes & Edges & \mean{Betweenness} & \median{P}\textsubscript{gene} & $\rho_{consensus}$\\
\hline
SConES GS & 5 & 0 & 9\,805 & 2.7 \texttimes{} 10\textsuperscript{-5} & 0.19\\
SConES GM & 28 & 2 & 4\,267 & 0.067 & 0.12\\
\end{tabular}
\begin{tablenotes}
\mean{Betweenness}: mean betweenness of the selected genes in the full network; \median{P}\textsubscript{gene}: median P-value of the selected genes; $\rho_{consensus}$: Pearson's correlation with the consensus network.
\end{tablenotes}
\end{threeparttable}
\end{table}

>>>>

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$, Christine Lonjou$^{1,2,3}$, Fabienne Lesueur$^{1,2,3}$,\\
GENESIS Study collaborators, Dominique Stoppa-Lyonnet$^{4,5,6}$,\\
Nadine Andrieu$^{1,2,3}$, Chloé-Agathe Azencott$^{3,1,2}$}
\date{$^{1}$Institut Curie, PSL Research University, F-75005 Paris, France;\\
  $^{2}$INSERM, U900, F-75005 Paris, France;\\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^{4}$Service de Génétique, Institut Curie, F-75005 Paris, France;\\
  $^{5}$INSERM, U830, F-75005 Paris, France;\\
  $^{6}$Université Paris Descartes.\\
}
\hypersetup{
  pdfauthor={Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, GENESIS Study collaborators, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott},
  pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
  pdfkeywords={},
  pdfsubject={},
  pdflang={English}}
\begin{document}

\onecolumn
\maketitle

\Large\textbf{Abstract}

Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks highly associated to this phenotype. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value = $7.8 \times 10^{-5}$ ) and more central than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (\emph{HSPA1A}, \emph{HSPA1B}, and \emph{HSPA1L}) or mitochondrial ribosomes (\emph{MRPS30}, \emph{MRPS31}, \emph{MRPS18B}). We also observed a general dysregulation in the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility.

@@

\begin{table}[htbp]
  \caption{\label{tab:method_comparison} Summary of the differences between the studied algorithms.}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Components & Input & Scoring & Reference\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & 1 & Summary & -log\textsubscript{10}(P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \item \textbf{\emph{Field}}: field in which the algorithm was developed.\\
      \item \textbf{\emph{Nodes}}: the type of network, either gene (protein-protein interaction network usually) or a SNP network.\\
      \item \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored.\\
      \item \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than being dense in high scores and connected.\\
      \item \textbf{\emph{Components}}: number of connected subnetworks in the solution.\\
      \item \textbf{\emph{Input}}: genotype data or GWAS summary statistics.\\
      \item \textbf{\emph{Scoring}}: how SNP/gene P-values are transformed into node scores.\\
      \item \textbf{\emph{Reference}}: original publication featuring the algorithm.
    \end{tablenotes}
  \end{threeparttable}
\end{table}

@@

\subsection{Network methods successfully identify genes associated with breast cancer}
\label{results:separate_networks}
\begin{table}[htbp]
  \centering
  \begin{threeparttable}
\caption{\label{tab:gene_solutions}
Summary statistics on the results of multiple network methods on the gene-gene interaction network. The first row contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrrrr}
Network & \# genes & \# edges & \mean{Betweenness} & \median{P}\textsubscript{gene} & \(\rho\)\textsubscript{consensus}\\
\hline
HINT HT & 13\,619 & 142\,541 & 16\,706 & 0.46 & 0.066\\
\hline
Consensus & 55 & 117 & 74\,062 & 0.0051 & 1\\
dmGWAS & 194 & 450 & 49\,115 & 0.19 & 0.41\\
heinz & 4 & 3 & 113\,633 & 0.0012 & 0.21\\
HotNet2 & 440 & 374 & 7\,739 & 0.048 & 0.31\\
LEAN & 0 & 0 & - & - & -\\
SConES GI & 0 (1) & 0 & - & - & -\\
SigMod & 142 & 249 & 92\,603 & 0.0083 & 0.73\\
\end{tabular}
\begin{tablenotes}
  \item \textbf{\emph{\# genes}}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses.
  \item \textbf{\mean{Betweenness}}: mean betweenness of the selected genes in the PPIN.
  \item \textbf{\median{P}\textsubscript{gene}}: median P-value of the selected genes.
  \item \textbf{\(\rho\)\textsubscript{consensus}}: Pearson's correlation between the subnetwork and the consensus network.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

\begin{table}[htbp]
\begin{threeparttable}
\caption{\label{tab:snp_solutions}
Summary statistics on the results of SConES on the three SNP-SNP interaction networks. The first row within each block contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrllr}
Network & SNPs & Edges & Subnetworks & \mean{Betweenness} & \median{P}\textsubscript{SNP}\\
\hline
GS & 197\,083 & 197\,060 & - & 2.03 \texttimes{} 10\textsuperscript{7} & 0.49\\
SConES GS & 1\,590 & 1\,585 & 5 & 2.52 \texttimes{} 10\textsuperscript{7} & 0.023\\
\hline
GM & 197\,083 & 6\,442\,446 & - & 3.99 \texttimes{} 10\textsuperscript{6} & 0.49\\
SConES GM & 1\,692 & 177\,611 & 5 & 4.40 \texttimes{} 10\textsuperscript{6} & 0.055\\
\hline
GI & 197\,083 & 28\,733\,720 & - & 1.46 \texttimes{} 10\textsuperscript{6} & 0.49\\
SConES GI & 408 & 539 & 5 & 9.33 \texttimes{} 10\textsuperscript{6} & 0.076\\
\end{tabular}
\begin{tablenotes}
\item \mean{Betweenness}: mean betweenness of the selected SNPs in the corresponding full network.
\item \median{P}\textsubscript{SNP}: median P-value of the selected SNPs.
\end{tablenotes}
\end{threeparttable}
\end{table}

\begin{table}[htbp]
\begin{threeparttable}
  \caption{\label{tab:scones_gene_solutions}
Summary statistics on the results of multiple network methods on the gene-gene interaction network. The first row contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrrlr}
Network & Genes & Edges & \mean{Betweenness} & \median{P}\textsubscript{gene} & $\rho_{consensus}$\\
\hline
SConES GS & 5 & 0 & 9\,805 & 2.7 \texttimes{} 10\textsuperscript{-5} & 0.19\\
SConES GM & 28 & 2 & 4\,267 & 0.067 & 0.12\\
\end{tabular}
\begin{tablenotes}
\item \mean{Betweenness}: mean betweenness of the selected genes in the full network.
\item \median{P}\textsubscript{gene}: median P-value of the selected genes; $\rho_{consensus}$: Pearson's correlation with the consensus network.
\end{tablenotes}
\end{threeparttable}
\end{table}
